Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-30 manu-draft.Rmd
all analyses now use appropriate summary estimates for non-detects #1, need to update text results

The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results. For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.

@@

```{r wqsumtab}
cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point. Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a. Summaries are median, maximum, and minimum values. Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
```

```{r wqcmptab}
cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.  Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
```

```{r mcrtab}
cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
totab <- mcrtab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savtab}
cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
totab <- savtab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

>>>>

The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results. Water quality values below detection were removed prior to PCA.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. Secchi observations that were visually identified on the bottom (12 observations of 518 total) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and can be evaluated with methods described below. Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.  These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from the NADA R package.  For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.  Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.  PCA and correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.

@@

```{r wqsumtab}
cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point. Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a. Summaries are median, maximum, and minimum values. Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.  The final column shows the percentage of total observations that were below detection limits (including Secchi on bottom).  Medians denoted by "-" could not be calculated due to insufficient values above detection.'
knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
```

```{r wqcmptab}
cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.  Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
```

```{r mcrtab}
cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrtab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savtab}
cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savtab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```
